Investigating Determinism of Gaming Engines for Autonomous Vehicle Verification

### 2020-03-13 paper/cav_determinisim_main.tex
added restricted results section and changed conclusion

\begin{table*}[b]
\centering
\begin{tabular}{clclccc}
\toprule
Test 		& Actors 						 & Collision & Collision Type 			 & $d_l$ (m) & $n$  &$\max\sigma$ (m) \\ \midrule
1       & Two vehicles                   & No  		 & N/A 			  			 & 2 		 & 1000 & 1000 \\
2       & Two vehicles                   & Yes 		 & Vehicle and Vehicle 		 & 2 		 & 1000 & 1000 \\
3       & Two vehicles and a pedestrian  & No  		 & N/A 			  			 & 2 		 & 1000 & 1000 \\
4       & Two vehicles and a pedestrian  & Yes 		 & Vehicle and Pedestrian 	 & 2 		 & 1000 & 1000 \\
5       & Two pedestrians                & No  		 & N/A 			  			 & 2 		 & 1000 & 1000 \\
6       & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 2 		 & 1000 & 1000 \\
% 6       & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 0.4 		 & 1000 & 1000 \\
% 8       & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 20 		 & 1000 & 1000 \\
\bottomrule
\end{tabular}
\caption{Test scenarios where $d_l$ is the \textit{look ahead distance}, $n$ is the number of repeats and $\max\sigma$ is the maximum simulation deviation. CONSIDER removing entries 6 and 8 if they do not contribute towards outcomes and remove references to $d_l$. \textbf{remove all reference to $d_l$}}
\label{TableOfExperiments}
\end{table*}

@@

\textbf{Abanoub: can you change the diagram to be a staggered t-junction as the DMR is just confusing for this example. Also can we zoom in on the lower right area. Tests 5 and 6 have different look ahead values, which is correct?}

@@

\begin{itemize}[leftmargin=*]
    \item System memory: \texttt{memtest86}\footnote{\url{https://www.memtest86.com/}} full suite of tests ran, all passed.
    \item Graphical memory: \texttt{cuda\_memtest}\footnote{see \url{https://github.com/ComputationalRadiationPhysics/cuda_memtest}} full details of tests found here~\cite{shi2009testing}, all tests passed.
    \item Non-uniform memory access: \texttt{numactl}\footnote{\url{https://linux.die.net/man/8/numactl}} \textbf{we need to run these test to look for $\max\sigma$ variations}
\end{itemize}

@@

\begin{figure}[h]
    \centering
    \begin{subfigure}{.24\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
        \caption{}
        \label{Test_a}
    \end{subfigure}
    \begin{subfigure}{.24\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
        \caption{}
        \label{Test_b}
    \end{subfigure}
    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}
\end{figure}

@@

% ***************************************************
%  RESULTS
% ***************************************************
\subsection{Results And Discussion}\label{ResultsSection}
The summary of all the tests are shown in the final column of Table~\ref{TableOfExperiments} where $\max\sigma$ is the maximum deviation across all resource utilisation levels. From these summary results it is clear that scenarios with only pedestrian actors not involved in collisions are those that are consistently deterministic over all resource utilisation levels \textbf{(is this true, any other cases?)}. However, for other scenarios involving vehicles or mixed actors resource utilisation level has a significant impact on \textit{simulation variance}.

@@

However, the non-deterministic results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are an overly pessimistic/conservative account of the situation. Examining these specific results (test 2, 4 \& 6??) as a function of simulation time reveals further information about the simulation variance before and after an actor collision callback. Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from deterministic prior to the vehicle collision and is non-deterministic afterwards. This pattern of deterministic behaviour prior to collision is maintained up to resource utilisation levels of approximately 50\%, Fig.~\ref{CarsCollisionPrePost}. This time series examination is repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}. \textbf{What is the summary data for this scenario?} Similarly to vehicle collisions, the variation of $\max\sigma$ for pedestrian collisions indicates deterministic behaviour below 95\% resource utilisation, Fig.~\ref{CarsPeopleCollisionPrePost}. This is a key finding of this work and suggests that verification engineers should consider this as an end condition for terminating tests as any results after this point are not reliable.

The second key finding of this work is illustrated in Fig.~\ref{CarsPeopleCollsionsCG25}. In this scenario (test 4), there is a collision between Car ID 2 (solid line) and Ped ID 3 (dot dash line) which occurs at a simulation time of approximately $6s$ and the other vehicle, Car ID 1, is not involved in the collision. There are three key observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays non-deterministic actor position post-collision as describe above. Secondly, that the pedestrian involved in the collision (Ped ID 3) displays deterministic behaviour even after the collision \textbf{is this because the walker is killed and removed from the game?}. And, thirdly we observed a delayed effect on Car ID 1 showing high simulation variance post-collision but with a $5s$ delayed effect \textit{even though this vehicle was not involved in the collision}. This final point should be of particular concern to verification engineers and research practitioners in the field as it implies that any collision between actors will affect the simulation variance of all actors in the simulation and could potentially result in false-positive assertion violations.

\textbf{Strange gap post-collision for few seconds}

\textbf{Should we plot a pre-collision only summary chart for all scenarios?}

\textbf{Should we rename 'test' to 'scenario???'}

@@

\begin{figure}[!t]
    \centering
    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV2.pdf}
    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels. Tests 5 and 6 overlap having almost identical results. \textbf{ Change to $cm$ level in the graphs. Move 0 stress level data to 5\% to reflect simulation resources required.}}
    \label{ExperimentsStressSummary}
\end{figure}

\begin{figure*}[h]
    \centering
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V2.pdf}
        \caption{}
        \label{CarsCollisionCG25}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV2.pdf}
        \caption{}
        \label{CarsCollisionPrePost}
    \end{subfigure}
    \caption{Vehicle to vehicle collision (test 2) showing (a) mean deviation against simulation time for 25\% resource utilisation and (b) mean deviation pre- and post-collision against resource utilisation. \textbf{Change Y axis value to $\max\sigma (m)$}}
\end{figure*}

\begin{figure*}[h]
    \centering
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25_V2.pdf}
        \caption{}
        \label{CarsPeopleCollsionsCG25}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV2.pdf}
        \caption{}
        \label{CarsPeopleCollisionPrePost}
    \end{subfigure}
    \caption{Vehicle to pedestrian collision (test 4) showing (a) mean deviation against simulation time for 25\% stress and (b) mean deviation pre- and post-collision for different resource utilisation. \textbf{Change Y axis value to $\max\sigma (m)$}}
\end{figure*}

@@

% ***************************************************
%  CONCLUSION
% ***************************************************
\section{Conclusions}\label{s:conclusion}
A detailed description of how practitioners in this field can resolve the non-deterministic operational domains of their simulation is presented. Potential sources of non-deterministic simulation behaviour are discussed and some of these are resolved, such as floating point error, and other are left unanswered such as scheduling policy. %
A case study was used to ascertain the boundaries of non-deterministic operation of the Carla driving simulator which seem to exist at collision events and when resource utilisation levels exceeded 50\% for the particular system used in this study. A collision is not necessarily a problem as long as the simulation is stopped as soon as it is detected as even actors not involved directly will behave non-deterministically.

>>>>

\begin{table*}[b]
\centering
\begin{tabular}{clclccc}
\toprule
Test & Actors 					 	  & Collision 	 & Collision Type 			 & $n$  & $\max\sigma$ (m) & $\max\sigma$ (m) \\ 
	 & 								  & 			 & 							 &	    & (unrestricted)  & (restricted)	\\ \midrule
1    & Two vehicles                   & No  		 & N/A 			  			 & 1000 & 0.03 & $7.0{\times}10^{-3}$ \\
2    & Two vehicles                   & Yes 		 & Vehicle and Vehicle 		 & 1000 & 0.31 & $9.8{\times}10^{-3}$ \\
3    & Two vehicles and a pedestrian  & No  		 & N/A 			  			 & 1000 & 0.07 & $5.2{\times}10^{-4}$ \\
4    & Two vehicles and a pedestrian  & Yes 		 & Vehicle and Pedestrian 	 & 1000 & 0.59 & $1.5{\times}10^{-12}$ \\
5    & Two pedestrians                & No  		 & N/A 			  			 & 1000 & $5.6{\times}10^{-13}$ & $5.6{\times}10^{-13}$ \\
6    & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 1000 & $5.6{\times}10^{-13}$ & $5.6{\times}10^{-13}$ \\
\bottomrule
\end{tabular}
\caption{A description of the test scenarios showing the test number, the actors included and if a collision occurred and between which actors. Where $n$ is the number of repeats and $\max\sigma$ is the maximum simulation deviation. The term \textit{unrestricted} refers to an unrestricted subset of the results and allowing system states of any resource utilisation whereas \textit{restricted} employs results filtering to pre-collision and limits system resource utilisation levels to 75\%.}
\label{TableOfExperiments}
\end{table*}

@@

@@

\begin{itemize}[leftmargin=*]
    \item System memory: \texttt{memtest86}\footnote{\url{https://www.memtest86.com/}} full suite of tests ran, all passed.
    \item Graphical memory: \texttt{cuda\_memtest}\footnote{see \url{https://github.com/ComputationalRadiationPhysics/cuda_memtest}} full details of tests found here~\cite{shi2009testing}, all tests passed.
    \item Non-uniform memory access: \texttt{numactl}\footnote{\url{https://linux.die.net/man/8/numactl}} was used to fix the simulator to a single core but no improvement in simulation variance was observed and not used for subsequent testing.
\end{itemize}

@@

\begin{figure}[!t]
    \centering
    \begin{subfigure}{.24\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
        \caption{}
        \label{Test_a}
    \end{subfigure}
    \begin{subfigure}{.24\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
        \caption{}
        \label{Test_b}
    \end{subfigure}
    \caption{Schematic of test scenarios for (a) tests 1-4, (b) tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}
\end{figure}

\begin{figure}[!t]
    \centering
    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV2.pdf}
    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels. Tests 5 and 6 overlap having almost identical results.}
    \label{ExperimentsStressSummary}
\end{figure}

@@

% ***************************************************
%  RESULTS
% ***************************************************
\subsection{Results And Discussion}\label{ResultsSection}
The summary of all the tests are shown in the final column of Table~\ref{TableOfExperiments} where $\max\sigma$ is the maximum deviation across all resource utilisation levels. From these summary results it is clear that scenarios with only pedestrian actors are those that are consistently deterministic over all resource utilisation levels. However, for other scenarios involving vehicles or a mixture of actors types resource utilisation level has a significant impact on \textit{simulation variance}.

@@

However, the non-deterministic results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation. Examining these specific results, tests 2 \& 4, in more detail as a function of simulation time reveals further information about the simulation variance before and after an actor collision. Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from deterministic prior to the vehicle collision and is non-deterministic post-collision. This pattern of deterministic behaviour prior to collision is maintained up to a resource utilisation level of approximately 75\%, Fig.~\ref{CarsCollisionPrePost}. This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}. Similarly to vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates pre-collision deterministic behaviour with up to 75\% resource utilisation, Fig.~\ref{CarsPeopleCollisionPrePost}. This is a key finding of this work and suggests that verification engineers should consider this as an end condition for terminating tests as any post-collision results are not reliable.

The second key finding of this work is illustrated in Fig.~\ref{CarsPeopleCollsionsCG25}. In this scenario (test 4), there is a collision between Car ID 2 (solid line) and Ped ID 3 (dot dash line) which occurs at a simulation time of approximately $6s$ and the other vehicle, Car ID 1, is \textit{not involved in the collision}. There are three key observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays non-deterministic actor position post-collision as describe above. Secondly, that the pedestrian involved in the collision (Ped ID 3) displays deterministic behaviour throughout the test\footnote{note that in Carla the pedestrian object is destroyed post-collision hence the flat line following the collision}. And, thirdly we observed a delayed effect on Car ID 1 showing high simulation variance with a $5s$ delayed effect \textit{even though this vehicle was not involved in the collision}. This final point should be of particular concern to verification engineers and research practitioners in the field as it implies that any collision between actors will affect the simulation variance of all actors in the simulation and could potentially result in false-positive assertion violations.

% \textbf{Strange gap post-collision for few seconds}

% \textbf{Should we rename 'test' to 'scenario???'}

@@

\begin{figure*}[h]
    \centering
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V3.pdf}
        \caption{}
        \label{CarsCollisionCG25}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV3.pdf}
        \caption{}
        \label{CarsCollisionPrePost}
    \end{subfigure}
    \caption{Vehicle to vehicle collision (test 2) showing (a) mean deviation against simulation time for 25\% resource utilisation and (b) mean deviation pre- and post-collision against resource utilisation.}
\end{figure*}

\begin{figure*}[h]
    \centering
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25_V3.pdf}
        \caption{}
        \label{CarsPeopleCollsionsCG25}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV3.pdf}
        \caption{}
        \label{CarsPeopleCollisionPrePost}
    \end{subfigure}
    \caption{Vehicle to pedestrian collision (test 4) showing (a) mean deviation against simulation time for 25\% stress and (b) mean deviation pre- and post-collision for different resource utilisation.}
\end{figure*}

@@

% ***************************************************
%  CONCLUSION
% ***************************************************
\section{Conclusions}\label{s:conclusion}
Practitioners in this field can resolve the non-deterministic operational domains of their simulation can be found by following the method outlined in this paper. Potential sources of non-deterministic simulation behaviour have been discussed such as floating point rounding error and scheduling policy. A case study was used to ascertain the boundary between deterministic and non-deterministic operation of the Carla driving simulator. Results suggest non-deterministic outcomes when resource utilisation levels exceeded 75\% or when vehicle collisions occur for the particular hardware used in this study. Furthermore it was found that all actors, even those not involved directly in a collision, will behave non-deterministically if a collision callback to the game engine is made. Best operating practice follows that the simulation should be stopped as soon as a collision callback is detected or resource utilisation levels exceed 75\% although this value may be system specific.
